What is an Ancestral Recombination Graph?

### 2022-03-03 paper.tex
Rephrase explanation of recombination nodes

\section*{Genome ARGs}
% We're not the first to talk about ARGs and pedigrees, and we're
% not doing anything controversial.
The deep connection between ARGs and pedigrees is of course
well known
\citep[e.g.][]{wakeley2012genetics,gusfield2014recombinatorics,
speed2015naturereviewsgenetics}
and recent discussions have illustrated the
relationship~\citep{mathieson2020ancestry,brandt2021evaluation}, % more??
showing how the coalescence and recombination events in an ARG
correspond to particular genomes in the pedigree.
As discussed in the previous section, the definition of an ARG
as a \emph{data structure} in terms of these two events is problematic,
fundamentally limiting the patterns of ancestry that can be described.
Following \cite{mathieson2020ancestry} instead of focussing on events,
we define the ARG data structure in terms \emph{genomes} and the patterns of
transmission between them.
% Bit of duplication here, but good to make the point about being
% motivated by real existing datasets not some academic desire to
% model complexity that isn't there.
This, coupled with
the explicit association of genome coordinates with edges, allows
us to represent even the bewilderingly complex patterns of relatedness
present in modern datasets.

@@

The focus on modelling genomes rather than events has some interesting
and important consequences. Firstly, the graph may contain nodes that
do not correspond to any particular event in the ancestry of the samples.
For example, node \textsf{H} is the direct ancestor of \textsf{D} in
Fig~\ref{fig-arg-in-pedigree}, and is not the product of either a
coalescence or a recombination. Such nodes would usually be removed
from the graph so that \textsf{D} descends directly from \textsf{K} (see
section XXX) but there is no necessity for this from a representational
perspective. This ability can be useful in practice and could be required if,
% Do we need to spell this out?
for example, we have a deeply sequenced pedigree of
haplodiploids. % e.g. Honeybees? Be great to cite an example!

Secondly, the way that we model recombination is subtly different.
Rather than having a single node that represents the \emph{event},
there are three nodes (representing the recombinant offspring and
two parental genomes) and two edges with associated genome coordinate
intervals (representing the precise inheritance patterns).
For example, in Fig~\ref{fig-arg-in-pedigree}, node \textsf{C} is the
recombinant descendent of \textsf{E} and \textsf{F}, inheriting positions
0 to 7 from $E$ and genome positions 7 to 10 from $F$. Any particular
node can be both a recombinant (it is the recombined product of
its parent's genomes) and a common ancestor (the ancestry from
multiple samples meet at this node; see the next section).
% Here was can talk about representing the recombination by a
% recombinant and 2 parent "recombination nodes" rather than
% a single "recombination node" as in the normal ARG representation.
% We may also wish to note that in this example the recombination nodes
% are shared between siblings, and can also be coalescent nodes
% this ties in with stuff discussed 3 paras below.
[More stuff here: clarify why we assign coordinates to edges
rather than nodes. Basically lay out all the stuff that we can
talk about \emph{without} going in to the details of
ancestry resolution.]

% Asking the reader to flick back to figure 2 here, not great.
The differences between EARGs
and GARGs are subtle but important, and illustrated in
Fig.~\ref{fig-arg-data-structure}, which shows a simple ARG
with three samples and one recombination.
Fig.~\ref{fig-arg-data-structure}A shows this topology as a
classical Griffiths-like EARG, and
Fig.~\ref{fig-arg-data-structure}B shows it as a GARG.
The two are almost identical, except we model the recombination
event via the two parental genomes in nodes $3$ and $4$ and explicitly
store the genomic interval that the child inherited from the
parent with every edge.
This approach may seem wasteful since we must store an extra
node for each recombination and genomic interval information
with every edge (even though many simply span the full genome).
However, asymptotically the amount of information stored is the
same, and as we see in the next section, keeping interval
information for all edges enables useful generalisations.
There is also much simplicity to be gained by having
a single type of node and by treating all edges in the same way.

>>>>

\section*{Genome ARGs}
% We're not the first to talk about ARGs and pedigrees, and we're
% not doing anything controversial.
The deep connection between ARGs and pedigrees is of course
well known
\citep[e.g.][]{wakeley2012genetics,gusfield2014recombinatorics,
speed2015naturereviewsgenetics}
and recent discussions have illustrated the
relationship~\citep{mathieson2020ancestry,brandt2021evaluation}, % more??
showing how the coalescence and recombination events in an ARG
correspond to particular genomes in the pedigree.
As discussed in the previous section, the definition of an ARG
as a \emph{data structure} in terms of these two events is problematic,
fundamentally limiting the patterns of ancestry that can be described.
Following \cite{mathieson2020ancestry} instead of focussing on events,
we define the ARG data structure in terms of \emph{genomes} and the
patterns of transmission between them.
% Bit of duplication here, but good to make the point about being
% motivated by real existing datasets not some academic desire to
% model complexity that isn't there.
This, coupled with
the explicit association of genome coordinates with edges, allows
us to represent even the bewilderingly complex patterns of relatedness
present in modern datasets.

@@

The focus on genomes rather than events has some interesting
and important consequences. Firstly, a gARG may contain nodes that
do not correspond to any particular event in the ancestry of the samples.
For example, node \textsf{H} is the direct ancestor of \textsf{D} in
Fig~\ref{fig-arg-in-pedigree}, and is not the product of either a
coalescence or a recombination. Such nodes would usually be removed
from the graph so that \textsf{D} descends directly from \textsf{K} (see
section XXX) but there is no necessity for this from a representational
perspective. This ability can be useful in practice and could be required if,
% Do we need to spell this out?
for example, we have a deeply sequenced pedigree of
haplodiploids. % e.g. Honeybees? Be great to cite an example!

Secondly, the natural way to represent the process of recombination in a
gARG is not as a single ``recombination node'' (as in an eARG), but with
two parent genomes between which recombination takes place (e.g.
nodes \textsf{E} and \textsf{F} in Fig~\ref{fig-arg-in-pedigree}), and a
single resulting ``recombinant'' genome (e.g. \textsf{C}). The
edges that link this recombinant with its two parents carry the information
concerning breakpoint location that would normally be present on the
"recombination node" in an eARG. For example, node \textsf{C} in
Fig~\ref{fig-arg-in-pedigree} inherits genome positions 0 to 7 from \textsf{E}
and positions 7 to 10 from \textsf{F}, indicating that the breakpoint is at position 7.
Unlike the eARG, the gARG need not make any distinction
between common ancestor and recombination nodes: indeed, a
genome can be simultaneously involved in recombination as well as being
a common ancestor, as seen most clearly in nodes \textsf{E} and \textsf{F}
in Fig~\ref{fig-arg-in-pedigree}b. Indeed, in this case, these nodes are involved in
 \emph{two} separate recombinations, giving rise to the genomes in two siblings.

The gARG method for representing recombinations may seem wasteful,
since more than one node is required for each recombination,
and genomic interval information must be stored on every edge.
However, asymptotically the amount of information stored is the
same, and there is much simplicity to be gained by having
a single type of node and by treating all edges in the same way.
More importantly --- as we see in the next section --- we gain the
ability to simplify the ARG by removing nodes without affecting the
local tree topologies, and also the ability to record only the
regions of genome relevant to the final, sample genomes. This last
feature turns out to be the key to efficient processing of the ARG.

[Possibly more stuff here: clarify why we assign coordinates to edges
rather than nodes. Basically lay out all the stuff that we can
talk about \emph{without} going in to the details of
ancestry resolution.]
